Framing Effects in the Wild

### 2022-08-31 paper/Manuscript_Framing.Rmd
Revise Text
- Add Discussion of Assumptions in the Translation of Experiments
- Move DPA-corrected estimates into main paper
- move raw treatment estimates into appendix
- Move Salience plot in Appendix
- Fix estimation section (drop reader effect)
- Improve Abstract
- Improve Intro
- Add section labels
- Improve Theory
- Add Figure Labels
- Improve Review of Existing Work
- Clarify Case Description

---
title: "Framing Effects in the Wild"
author: "Nicolai Berk"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: false
    fig_caption: yes
    includes:
      in_header: preamble.tex
# subtitle: Manuscript prepared for the 12th Annual Conference of the European Political Science Association in Prague, Czech Republic
fontsize: 12pt
geometry: margin=1.5in
abstract: "Does news framing affect public opinion? While a large body of experimental work suggests that individual attitudes can be easily altered by emphasizing different aspects of an issue, little is known about how these studies translate into the real world. I exploit a rare natural change of news framing in the major German tabloid newspaper *Bild* to study whether experimental estimates of framing effects translate into the real world. In a first step, I apply supervised transformer models to a novel dataset of 2.5M news articles and provide evidence of a substantial shift in the paper's migration framing following an editorial change. Second, I employ a difference-in-differences design using panel data from over 16,000 individuals in ten waves to assess the precise impact on migration attitudes. The findings show that, despite a major change in migration framing, immigration attitudes and related variables remain unaffected. News framing seems to affect attitudes only within limited windows of opportunity early in the politicization of political issues."
bibliography: C:/Users/nicol/Dropbox/Studium/BibTex/PhD.bib
nocite: |
  @CologneReport2016
---

@@

*I am thankful for extensive feedback and other support on this manuscript from Christian Arnold, Daniel Bischof, Philipp Broniecki, Jóhanna Ýr Bjarnadóttir, Elias Dinas, Florian Foos, Violetta Haas, Felix Hagemann, Julian Heid, Christina Gahn, Heike Klüver, Martin Vinæs Larsen, Matthew Levendusky, Will Lowe, Thomas M. Meyer, Rune Slothuus, Judith Spirig, Antonio Valentim as well as participants in the EPSIP Colloquium at HU Berlin, CIVICA EGN Workshop at the EUI, CompText Conference 2022, and the Political Behaviour Section at Aarhus University. All remaining confusions are my own.*

<!--# I also thank the two RAs, Robin Groß and Nathalie Aerts, for annotating the news articles.  -->

@@

Does news framing affect public opinion? While a large body of experimental evidence suggests that the emphasis of a certain consideration can substantially alter issue attitudes, we know little about the translation of these findings into real-world environments. Prior work has shown that the external validity of such experiments might be seriously complicated by environments of diverse frames [@Chong2013] and actors [@Slothuus2010], as well as the seeking and avoidance of specific content [@Lau2021a; @Taber2006]. Especially on contested issues, citizens might be more inclined to resist persuasion [@Arceneaux2009]. Contemporary news environments provide consumers with especially diverse content and maximize their abilities to choose specific outlets and avoid others [@Bennett2008], thereby increasing the gap between the real-world environment and experimental studies. Additionally, the very nature of experiments is prone to over-estimate effects in the real world [@Barabas2010]. This is the reason why, fifteen years ago, Donald Kinder called for framing research to extend its toolkit: "[E]xperimental results can always be questioned on their generalizability, and framing effects are no exception. [...] A more balanced reading of frame effects requires methodological diversification, experiments and studies oriented to the world outside" [-@Kinder2007, 157; see also @Busby2019; @Leeper2020].

I provide the - to my knowledge - first observational estimate of framing effects by exploiting a rare shift in the migration framing of Germany's largest tabloid newspaper, *Bild*. The case is ideal to assess the everyday influence of issue frames on public opinion, as one outlet substantially changed its framing while other papers maintained theirs. By employing BERT models for the measurement of crime frames in a corpus of 2.5 million newspaper articles, I am able to clearly identify the treatment - something that is usually left to experimental studies. The fielding of the German Longitudinal Election Study in the same period provides the rare opportunity to precisely assess the effect of this changing framing on readers' migration attitudes in a *difference-in-differences (DiD) design*.

[might want to mention frames in mind and frames in communication distinction here and how it informs research design (follow Rune & Martin Bisgaard)]

The results indicate that even a strong increase in the use of crime frames did not affect consumers' immigration attitudes. The estimated effect is statistically indistinguishable from zero and significantly smaller than even minimal effect magnitudes. Related issue attitudes, voting behaviour, and issue importance are similarly unaffected, as is the association of immigration attitudes with the frame provided. Additional evidence implies that immigration attitudes had crystallized at the time of the shift, such that individuals were immune to changing framing.

Overall, these results fundamentally challenge the external validity of framing effects research and suggest that the influence of individual news outlets on public opinion is more limited than previously thought. Even a strong change in news framing did not affect political attitudes. Instead, it seems that timing is a crucial factor in the effectiveness of news frames.

# Framing Effects on Public Opinion

"Framing is the process by which a communication source, such as a news organisation, defines and constructs a political issue or public controversy" [@Nelson1997Rally]. Different forms of frames have been conceptualized. When I talk about 'frames' here, I refer to *emphasis frames*. These frames emphasize certain topics in relation to an issue, guiding the recipient to think about the issue with those considerations in mind that are promoted by the frame [@Leeper2020, 152]. For example, issues of increasing welfare contributions might be discussed with reference to inequality and providing chances to the poor, or by mentioning that higher welfare benefits might result in higher taxes. Individuals' support of a statement will differ dependent on the frame presented [@sniderman2004structure]. Many other forms of framing exist, such as equivalence frames [@Kahneman1979prospect], episodic frames [@Iyengar2005], or generic frames [@DeVreese2005]. The focus on emphasis frames in this study is owed to the large body of literature and generalized theory regarding their effect on attitudes about any issue (as opposed to generic and episodic frames), as well as their prevalence in the news media (which is not the case for equivalence frames).

Work using the emphasis framing concept of framing mostly builds on the expectancy-value model developed by @Ajzen2000. This model suggests that an attitude on a given issue is a function of two things: the value of considerations relating to that issue and the strength of their association with the issue. The evaluation of the issue is subdivided into a number of considerations [a "frame in mind"; @Chong2007a, 105; @Nelson1996], which will carry clear evaluations. For example, to form an opinion about whether immigration to one's country should be restricted, a person might consider the humanitarian conditions in the countries of origin (negative consideration regarding restrictive migration policy), as well as the risk of increased crime (positive consideration; note that the model does not differentiate whether a consideration is grounded in objective reality). Individuals weigh each of these considerations to form an overall opinion on the subject.

Figure \ref{fig:cogStor} visualizes this logic. A given issue (bottom square) carries an attitude $y$. The issue is associated with a number of different considerations (top circles), each of which carries an evaluation $x_i$, e.g. crime is viewed very negatively. Based on the strength of the association $w_i$ of a given consideration with the issue, a given consideration's evaluation is more or less reflected in the issue evaluation $y$, which is equal to the weighted sum of considerations. For example, if a given individual associates immigration strongly with an increase in criminal activity, welfare fraud, but only very weakly with humanitarian considerations, cultural enrichment, liberty, or economic growth, they will form a rather negative attitude about the issue, as the former factor more prominently in their reasoning about migration.

@@

How does this model incorporate news content and other forms of political communication? News content can emphasize different considerations of an issue [so called "frames in communication" @Nelson1996]. By being exposed to the consideration, an individual's cognitive association of the issue at hand with the raised consideration strengthens (or in more technical terms: the weight of the consideration $w_i$ increases). As a result, this specific consideration factors more prominently in the formulation of an overall attitude: "media frames influence opinions by stressing specific values, facts, or other considerations, endowing them with greater apparent relevance to the issue than they might appear to have under an alternative frame" [@Nelson1997Rally, 569].

@@

# Limits of Framing Experiments

Research on framing has shown for decades that citizens' attitudes and their issue associations are highly reactive to minor changes in the emphasis of specific considerations [@Chong2007b; @Chong2010; @Druckman2012Bias; @Kinder1990; @Nelson1996; @Nelson1997Rally; @sniderman2004structure; for reviews see @Busby2019; @Druckman2001; @Leeper2020]. However, it remains unclear how well these experimental findings translate into the real world. First, taking experimental effects as estimates of the real-world impact of framing means making strong assumptions about citizens' exposure to frames: while framing studies usually simply present a news article or other forms of political content to respondents, news content in the real world competes with other news and entertainment content to reach its audience. Changes in news framing itself might affect how likely readers are to consume news content: readers might be less likely to consume frames challenging their existing attitudes and reject news outlets containing such frames (and vice versa seek out attitudinally congruent information). This idea of 'selective exposure/avoidance' [@Bennett2008], or a 'confirmation bias' [@Taber2006] brings with it a major endogeneity problem making purely correlational evidence of framing effects rather uninformative.

Second, even if one assumes that the consumers receive the content, (e.g. buy a newspaper, visit a website, ...), it is not clear whether the content is actually consumed. For example, a reader might well buy a newspaper but skip an article after reading the headline because the topic is not of interest or the frame is at odds with their existing attitude [@Lau2021a]. As most framing studies do not offer choice of content to respondents, engagement remains an assumption when taking experimental studies as evidence of framing effects in the real world (for an exception see @Druckman2012Bias). Additionally, information from different sources, such as TV, social media, blogs, newspapers, ..., might 'crowd out' the framing of a single outlet: By consuming content from many sources, individuals might give less attention to a single outlet, thereby severely limiting the overall impact of that outlet's coverage.

Third, even when consuming the content, cognitive engagement with it might strongly differ in the real world compared to experimental contexts. Motivated reasoners presented with a plethora of news content about an issue will spend less time and cognitive energy on counter-attitudinal arguments than those supporting their viewpoint, lending more importance to considerations raised by amenable content [@Taber2006]. Experimental studies presenting only one frame per issue cannot take this bias into account. Consumption might also result in better or worse comprehension in the real world because respondents might pay more or less attention to real news content compared to experimental treatments [@Barabas2010]. Furthermore, there might be differences in how easily this translates into changing cognitive associations, given that those exposed might take real-world content more seriously. Additionally, the relative timing at which a frame is received, relative to other frames and the salience of the issue, affects the way in which individuals react to framing. Those citizens who received a different frame earlier might already have formed a crystallized opinion about the issue and thus reject subsequent attempts of persuasion [@Bechtel2015; @Druckman2012Bias; @Lecheler2009b].

To summarize, there are many reasons to assume that framing might have very different consequences in real-world environments than in their experimental emulations, and the above list is by no means comprehensive. When we take the experimental evidence as an approximation of the effectiveness of framing in the real world, we make many strong assumptions about citizens' exposure to, consumption of, and engagement with the frames in news content. Experiments can only incorporate few characteristics of the real world at a time. In order to test the external validity of framing effect studies, it is necessary to study observational data.

## Existing Observational Studies of Framing Effects

Despite repeated calls for the assessment of framing effects beyond experimental studies [@Busby2019; @Kinder2007; @Leeper2020], there is little research addressing this concern. While a large body of observational media effects research exists, it has not paid specific interest to framing, instead assessing the impact of changing news exposure broadly [@Bail2018; @Broockman2022; @Foos2022; @Grossman2022; @Guess2021; @Levendusky2022; @Spirig2020].

The precious small body of existing observational research on framing effects is often riddled with endogeneity issues. For example, @Jerit2008 investigates the effects of news framing on public opinion. She finds that framing is less effective in changing public opinion than engagement with opposing considerations, a finding that suggests that framing is rather ineffective. However, her study restricts itself to the assessment of correlational evidence. This is especially problematic regarding framing effects, as news content can affect consumers' inclination to consume news (see preceding section), and political actors might react to public opinion, thereby creating endogeneity issues. Others show with similar designs that only specific frames seem to drive issue importance [@Mclaren2018], but do not assess effects on attitudes.

To my knowledge, there are only two observational studies assessing the causal effect of exposure to specific content on political attitudes. @Slothuus2010b takes advantage of a sudden shift in a political parties' framing of retirement benefits, and shows that this changed framing has substantial impact on party supporters' attitudes about the issue. However, it is unclear how these findings translate to news framing effects. Given that - in this case - a positional change coincided with the changing framing of the issue and parties' special role in citizens' information process, it is impossible to distinguish the effect of partisan cue-taking from the persuasive effect of the frame itself.

The only study which assesses the precise impact of changing news content on political attitudes comes from @Djourelova2020. She exploits a ban of the term "illegal migrant" in the *Associated Press* and shows that this language has modest effects on migration attitudes. While the study itself reveals impressively how language can shape political attitudes, it constitutes a highly specific case and a rather minor change in the framing of migration. To gain a deeper understanding of how specifically framing affects immigration attitudes, it is necessary to assess broader changes in the representation of the immigration issue.

In sum, we lack an understanding of how framing affects political opinions in the real world. While experimental evidence suggests rather strong effects, it is unclear how these findings translate to the outside world. Observational evidence on media effects is abundant but has not precisely studied the impact of framing. The present study therefor exploits a naturally occurring change in the news framing of a particular issue *within* a specific outlet to isolate the effects of news framing on political opinions.

@@

@@

In order to test whether changing news content affected political attitudes, I exploit an editorial change in the largest German newspaper, *Bild*. As I will show, this editorial change resulted in a substantial shift in the papers' migration framing and is hence ideal to assess the effect of within-newspaper content shifts. The fielding of a large-scale panel survey in the same period allow to assess the impact of news content on migration attitudes.

In the summer of 2015, Germany became sanctuary for around 800,000 refugees who had fled war, violence, and famine in Syria and elsewhere. The German public debated how to deal with the newly arriving and for the first time since the post-war years, a radical right party was likely to win representation in the national parliament. Surprisingly, given its traditional populist style and right-wing takes on migration, the major tabloid newspaper *Bild* promoted an open stance towards the newly arriving in this period. This was often connected to chief editor Kai Diekmann, who hosted a refugee family in his home [@BR2015Diekmann].

In December 2016, Axel Springer, the company owning Bild, announced that Diekmann will leave the company. Less than a week after, it became public that prosecutors investigate Diekmann for sexual harassment. While the company claimed that this was not the reason for Diekmann's departure, the process might have been sped up as a result of the investigations\footnote{According to a former editor of the newspaper I talked to, rumors in the company at the time support this interpretation of events.} [@Spiegel2017Diekmann]. The migration coverage in *Bild* severely changed when Diekmann left the newspaper in February 2017 and Julian Reichelt took over the position as editor-in-chief. The paper increasingly started painting a picture of refugees as criminals on the tabloid's front page [@Niggemeyer2018Bild; @DLF2018Reichelt]. Headlines like "I killed 40 people and want asylum" increasingly dominated the migration coverage of the newspaper. In a speech months before he took over the position as chief editor, Reichelt outlined his motivation for such a shift in migration coverage: "I can assure you: nothing has hurt [*Bild*] economically as much as our clear, humane, empathetic stance in the refugee crisis" \footnote{Reichelt in a speech at Deutschlandfunk's conference "Formate des Politischen 2016": \url{https://vimeo.com/190347766}}. While he expressed his support for the decision to take this stance at the time, he argued that the German media shows too little tolerance towards anti-immigration attitudes and therefore excludes those holding such views from political discourse.

This case presents a unique case to study the effects of media content on public opinion. As will be shown below, a shift in migration coverage can be clearly identified. The change of editor itself was not discussed in the media, which means any opinion changes among readers are the result of the changing content [@Spirig2020]. Additionally, the timing of the editorial change is potentially exogenous, given the pressure from allegations against Diekmann. In summary, *this case is as close as possible to a field experiment in which a specific consideration of migration coverage is amplified in the coverage of one newspaper, but not others*.

[Outline research design, measurement of frames in communication and thought, following RUne and Martin Bisgaard]

@@

To measure immigration framing in the news, I collected 2.5 million newspaper articles from the websites of the most important German broadsheets \textit{Frankfurter Allgemeine Zeitung (FAZ), Spiegel Online (SPON), Süddeutsche Zeitung (SZ), Die Tageszeitung (TAZ), Die Welt} and the major German tabloid \textit{Bild} for the period 2013-2019 (for the market reach of each outlet, see table 1). Following the value-expectancy model, I identify crime frames in migration coverage in a two-step process: first, I use supervised machine learning to identify content about migration. Second, I identify content about crime *within* migration coverage by applying a similar supervised classifier. This approach closely resembles the value-expectancy model, as I first identify coverage about the *issue*, before identifying the considerations emphasized in the discussion of this issue.

To identify whether an article is about migration, I employ supervised machine learning. As migration content is rare, and training data needs to be somewhat balanced, articles cannot be selected randomly for annotation[^1]. Instead, I first construct a dictionary of terms related to migration. I use dictionary extension based on German GloVe word-embeddings\footnote{Downloadable from \url{https://deepset.ai/german-word-embeddings}.} to construct a comprehensive dictionary, and apply it to the articles. Based on the relative share of migration words in an article, I draw a stratified sample of 300 articles (100 from articles with no migration terms, 100 from the quarter with most migration terms and 100 from the remainder in between) for each newspaper for a total of 1,800 papers. Then, a coder hand-coded these articles, assessing whether their main topic is related to migration.

[^1]: If migration content made up about 3% of all articles and the sample were selected randomly, one would find around 30 migration articles per 1000 articles, which provides rather limited information about the class of interest and results in worse recall, as the best guess in highly imbalanced data is simply the most common outcome.

@@

TreatmentTrendPlot() + 
  TreatmentDiDPlot()

@@

>>>>

---
title: "Framing Effects in the Wild"
author: "Nicolai Berk"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: false
    fig_caption: yes
    includes:
      in_header: preamble.tex
# subtitle: Manuscript prepared for the 12th Annual Conference of the European Political Science Association in Prague, Czech Republic
fontsize: 12pt
geometry: margin=1.5in
abstract: "Does news framing affect public opinion? While a large body of experimental work suggests that individual attitudes can be easily altered by emphasizing different aspects of an issue, little is known about how these studies translate into the real world. I exploit a rare change in the migration framing of the major German tabloid *Bild* to study whether experimental estimates of framing effects translate into the real world. In a first step, I develop a framework for the measurement of emphasis frames in political communication using supervised machine learning. Using this framework, I show that the emphasis of crime in *Bild*'s migration coverage severely increased. Second, I employ a difference-in-differences design using panel data from over 16,000 individuals in ten waves to assess the precise impact of this shift on migration attitudes. The findings show that, despite a major change in migration framing, a clear null effect on immigration attitudes and related variables can be observed. Tracing three key assumptions underlying the translation of experimental effect estimates suggests that news framing affects attitudes only within limited windows of opportunity early in the politicization of political issues."
bibliography: C:/Users/nicol/Dropbox/Studium/BibTex/PhD.bib
nocite: |
  @CologneReport2016
---

@@

*I am thankful for extensive feedback and other support on this manuscript from Christian Arnold, Daniel Bischof, Philipp Broniecki, Jóhanna Ýr Bjarnadóttir, Elias Dinas, Florian Foos, Violetta Haas, Felix Hagemann, Julian Heid, Christina Gahn, Heike Klüver, Martin Vinæs Larsen, Matthew Levendusky, Will Lowe, Thomas M. Meyer, Rune Slothuus, Judith Spirig, Antonio Valentim as well as participants of CompText 2022, EPSA 2022, ICA 2022, EPSIP Colloquium at HU Berlin, CIVICA EGN Workshop at the EUI,  and the Political Behaviour Section at Aarhus University. I also thank the two RAs, Robin Groß and Nathalie Aerts, for annotating the news articles. All remaining confusions are my own.*

@@

Does news framing affect public opinion? While a large body of experimental evidence suggests that the emphasis of a certain consideration can substantially alter issue attitudes, we know little about the translation of these findings into real-world environments. Prior work has shown that the external validity of such experiments might be seriously complicated by the current salience and politicisation of the issue of interest [@Arceneaux2009; @Bechtel2015; @Lecheler2009a], the availability of different frames [@Chong2007; @Chong2013; @Druckman2003; @Druckman2012] and communicating actors [@Slothuus2010], as well as the seeking and avoidance of specific content [@Lau2021a; @Taber2006]. Contemporary news environments provide consumers with especially diverse content and maximize their abilities to choose specific outlets and avoid others [@Bennett2008], thereby increasing the gap between the real-world environment and experimental studies. Additionally, the very nature of experiments is prone to over-estimate effects in the real world [@Barabas2010]. This is the reason why, fifteen years ago, Donald Kinder called for framing research to extend its toolkit: "[E]xperimental results can always be questioned on their generalizability, and framing effects are no exception. [...] A more balanced reading of frame effects requires methodological diversification, experiments and studies oriented to the world outside" [-@Kinder2007, 157].

In this article, I provide the - to my knowledge - first observational estimate of framing effects by exploiting a rare shift in the framing of a highly politicized issue - migration - in Germany's largest tabloid newspaper, *Bild*. The case is ideal to assess the everyday influence of issue frames on public opinion, as one outlet substantially changed its framing while other papers maintained theirs. Providing a novel framework for the measurement of emphasis frames using BERT models, I study migration framing in a corpus of 2.5 million newspaper articles. I show that the newspaper's migration coverage saw a strong increase in the use of crime frames after the editor in chief was replaced. The fielding of the German Longitudinal Election Study in the same period provides the rare opportunity to precisely assess the effect of framing on political attitudes using a *difference-in-differences (DiD) design*.

The results indicate that even a strong increase in the use of crime frames did not affect consumers' immigration attitudes. The estimated effect is statistically indistinguishable from zero and significantly smaller than even minimally relevant effect magnitudes. Related issue attitudes, voting behavior, and issue importance are similarly unaffected, as is the association of immigration attitudes with the frame provided. Tracing different assumptions underlying the translation of experimental estimates of framing effects into the real world implies that this lack in responsiveness is not an outcome of readers abandonment of the newspaper, nor the "crowding out" of information through other sources, but rather the crystallization of immigration attitudes prior to the change in framing.

Overall, these findings fundamentally challenge the external validity of a large body of framing effects research and suggest that the influence of individual news outlets on public opinion is more limited than previously thought. Even a strong change in news framing did not affect political attitudes. Instead, it seems that the news media faces limited windows of opportunity to shape public opinion during the politicization of novel political issues.

# Framing Effects on Public Opinion {#framing-effects}

"Framing is the process by which a communication source, such as a news organisation, defines and constructs a political issue or public controversy" [@Nelson1997Rally]. Different forms of frames have been conceptualized. When I talk about 'frames' here, I refer to *emphasis frames*. These frames emphasize certain topics in relation to an issue, guiding the recipient to think about the issue with those considerations in mind that are promoted by the frame [@Leeper2020, 152]. Many other forms of framing exist, such as equivalence frames [@Kahneman1979prospect], episodic frames [@Iyengar2005], or generic frames [@DeVreese2005]. The focus on emphasis frames in this study is owed to the generalized theory regarding their effect on attitudes about any issue (as opposed to generic and episodic frames), as well as their common use in news content (which is not the case for equivalence frames).

Work using the emphasis framing concept of framing mostly builds on the expectancy-value model developed by @Ajzen2000. This model suggests that an attitude on a given issue is a function of two things: the value of considerations relating to that issue and the strength of their association with the issue. The evaluation of the issue is subdivided into a number of considerations, which will carry clear evaluations. For example, to form an opinion about whether immigration to one's country should be restricted, a person might consider the humanitarian conditions in the countries of origin (negative consideration regarding restrictive migration policy), as well as the risk of increased crime (positive consideration; note that the model does not differentiate whether a consideration is grounded in objective reality). Individuals weigh each of these considerations (so-called "frames in thought"; @Druckman2001) to form an overall opinion on the subject.

Figure \ref{fig:cogStor} visualizes this logic. A given issue (bottom square) carries an attitude $y$. The issue is associated with a number of different considerations (top circles), each of which carries an evaluation $x_i$, e.g. crime is viewed very negatively. Based on the strength of the association $w_i$ of a given consideration with the issue, a given consideration's evaluation is more or less reflected in the overall attitude $y$, which is equal to the weighted sum of considerations. For example, if a given individual associates immigration strongly with an increase in criminal activity, welfare fraud, but only very weakly with humanitarian considerations, cultural enrichment, liberty, or economic growth, they will form a rather negative attitude about the issue, as the former factor more prominently in their reasoning about migration.

@@

How does this model incorporate news content and other forms of political communication? News content can emphasize different considerations of an issue (so-called "frames in communication"). By being exposed to the consideration, an individual's cognitive association of the issue at hand with the raised consideration strengthens (or in more technical terms: the weight of the consideration $w_i$ increases). As a result, this specific consideration factors more prominently in the formulation of an overall attitude: "media frames influence opinions by stressing specific values, facts, or other considerations, endowing them with greater apparent relevance to the issue than they might appear to have under an alternative frame" [@Nelson1997Rally, 569].

@@

# Limits of Framing Experiments {#framing-limits}

Research on framing has shown for decades that citizens' attitudes and their issue associations are highly reactive to minor changes in the emphasis of specific considerations [@Chong2007b; @Chong2010; @Druckman2012Bias; @Kinder1990; @Nelson1996; @Nelson1997Rally; @sniderman2004structure; for reviews see @Busby2019; @Druckman2001; @Leeper2020]. However, it remains unclear how well these experimental findings translate into the real world. Below, I outline three key assumptions in the translation of experimental estimates.

First, taking experimental effects as estimates of the real-world impact of framing means making strong assumptions about citizens' exposure to frames: while framing studies usually simply present news articles or other forms of political content to respondents, news content in the real world competes with other news and entertainment content to reach its audience. Changes in news framing itself might affect how likely readers are to consume news content: readers might be less likely to consume frames challenging their existing attitudes and reject news content containing such frames (and vice versa seek out attitudinally congruent information). This idea of 'selective exposure/avoidance' [@Bennett2008], or a 'confirmation bias' [@Taber2006] brings with it a major endogeneity problem making purely correlational evidence of framing effects rather uninformative. As most framing studies do not offer choice of content to respondents, engagement remains an assumption when taking experimental studies as evidence of framing effects in the real world [for an exception see @Druckman2012Bias].

Second, even if consumers receive the content, it is not clear how much an individual relies on this source when forming an opinion on the subject.  Information from different sources - such as TV, newspapers, social media, or alternative online media - might 'crowd out' the framing of a single outlet: by consuming content from many sources, individuals might give less attention to a given source, thereby severely limiting the overall impact of that outlet's coverage. Experimental evidence suggests that, when an individual receives competing frames, these cancel each other out, although the precise impact is dependent on the relative timing of exposure, the quality of the frame, as well as individual characteristics [@Chong2007b; @Chong2010; @Chong2013].

Third, cognitive engagement with frames might strongly differ in the real world compared to experimental contexts. Once individuals formed political attitudes, they might resist framing: Motivated reasoners presented with a plethora of news content about an issue will spend less time and cognitive energy on counter-attitudinal arguments than those supporting their viewpoint, lending more importance to considerations raised by amenable content [@Taber2006]. Experimental evidence suggests similar dynamics specifically for framing: when individuals are free to choose content, initial frames dominate the opinion-formation process, while novel frames are rejected [@Druckman2012Bias] and even serve to strengthen pre-existing attitudes [@Bechtel2015]. This also explains why citizens are less reactive to framing attempts on salient and politicized issues [@Lecheler2009b].

In sum, there are many reasons to assume that framing might have very different consequences in real-world environments than in their experimental emulations. If one takes experimental evidence as an approximation of the effectiveness of framing in the real world, they make many strong assumptions about citizens' exposure to, reliance on, and cognitive engagement with the frames in news content. Experiments can only incorporate few characteristics of the real world at a time. In order to test the external validity of framing effect studies, it is necessary to study observational data.

# Existing Observational Studies of Framing Effects

Despite repeated calls for the assessment of framing effects beyond experimental studies [@Busby2019; @Kinder2007; @Leeper2020], there is little research addressing this concern. While a large body of observational media effects research exists, it has not paid specific interest to framing, instead assessing the impact of changing news exposure on issue attitudes and voting behavior more broadly. Most of these studies are concerned with exposure to a different outlet [@Bail2018; @Broockman2022; @Foos2022; @Grossman2022; @Guess2021]. This is problematic, as citizens are often aware of outlets' political slant and engage with the coverage differently [@Baum2008; see also @Spirig2020 for a discussion of this issue]. The existing studies focusing on within-outlet changes in coverage do not identify changing framing directly and assess effects on a broad set of dependent variables, rather than specific issue attitudes [@Levendusky2022; @Spirig2020].

The precious small body of existing observational research specifically about framing is often riddled with endogeneity issues. For example, @Jerit2008 investigates the effects of news framing on public opinion. She finds that framing is less effective in affecting public opinion compared to engagement with opposing considerations, a finding that suggests that framing is rather ineffective. Others show with similar designs that only specific frames seem to drive issue importance [@Mclaren2018], but do not assess effects on attitudes. However, these studies restrict themselves to the assessment of correlational evidence. This is especially problematic regarding framing effects, as news content can affect consumers' inclination to consume news (see preceding section), and political actors might react to public opinion. As both variables affect each other at the same time (endogeneity), it is impossible to assess which direction causality runs.

To my knowledge, there are only two observational studies assessing the causal effect of exposure to specific content on political attitudes. @Slothuus2010b takes advantage of a sudden shift in a political parties' framing of retirement benefits, and shows that this changed framing has substantial impact on party supporters' attitudes about the issue. However, given that a positional change coincided with the changing framing of the issue and parties' special role in citizens' information process, one cannot distinguish the effect of partisan cue-taking from the persuasive effect of the frame itself. Additionally, it is unclear how these findings translate from parties' communication to news framing effects.

The only study which - to my knowledge - assesses the precise impact of changing news content on political attitudes comes from @Djourelova2020. She exploits a ban of the term "illegal migrant" in the *Associated Press* and shows that this language has modest effects on migration attitudes. While the study itself reveals impressively how language can shape political attitudes, it constitutes a highly specific case and a rather minor change in the framing of migration. To gain a deeper understanding of how specifically *framing* affects immigration attitudes, it is necessary to assess broader changes in the representation of the immigration issue.

It becomes obvious that the discipline lacks an understanding of how news framing affects political attitudes in the real world. While experimental studies find strong effects, several assumptions have to be made to translate them beyond the lab or survey context. Observational evidence on media effects has not precisely studied the impact of framing. To do so, it is necessary to assess within-outlet changes in news framing and their effect on specific issue attitudes using a design able to identify causal effects. The following section therefore presents a research design building on a rare change in the migration framing in Germany's largest tabloid, *Bild*. It then develops a framework to measure emphasis framing in political communication, and presents a panel dataset tracking migration attitudes in the time-frame of interest, as well as a *difference-in-difference* design to estimate the causal effect of the changing framing. The final section discusses the expected effect size from experimental studies, in order to provide a baseline for comparison.

@@

In order to test whether changing news content affected political attitudes, I exploit an editorial change in the largest German newspaper, *Bild*. As I will show, this editorial change resulted in a substantial shift in the papers' migration framing and is hence ideal to assess the effect of within-newspaper content shifts. The fielding of a large-scale panel survey in the same period allows to assess the precise impact of this change in news content on consumers' migration attitudes.

@@

In the summer of 2015, Germany became sanctuary for around 800,000 refugees who had fled war, violence, and famine in Syria and elsewhere. The German public debated how to deal with the newly arriving and for the first time since the post-war years, a radical right party was likely to win representation in the national parliament. Surprisingly, given its traditional populist style and right-wing takes on migration, the major tabloid newspaper *Bild* initially maintained a somewhat balanced coverage of the newly arriving, which - although portraying migrants more negatively than other outlets - even started a campaign called "#RefugeesWelcome" to gather support for the newly arriving [@Maurer2019; see @BildRefugeesWelcome]. This was often connected to chief editor Kai Diekmann, who hosted a refugee family in his home [@BR2015Diekmann].

In December 2016, Axel Springer, the company owning Bild, announced that Diekmann will leave the company. Less than a week after, it became public that prosecutors investigate Diekmann for sexual harassment. While the company claimed that this was not the reason for Diekmann's departure, the process might have been sped up as a result of the investigations\footnote{According to a former editor of the newspaper I talked to, rumors in the company at the time support this interpretation of events.} [@Spiegel2017Diekmann]. The migration coverage in *Bild* severely changed when Diekmann left the newspaper in February 2017 and Julian Reichelt took over the position as editor-in-chief. The paper increasingly started painting a picture of refugees as criminals on the tabloid's front page [@Niggemeyer2018Bild; @DLF2018Reichelt]. Headlines like "I killed 40 people and want asylum" increasingly dominated the migration coverage of the newspaper. In a speech months before he took over the position as chief editor, Reichelt outlined his motivation for such a shift in migration coverage: "I can assure you: nothing has hurt [*Bild*] economically as much as our clear, humane, empathetic stance in the refugee crisis" \footnote{Reichelt in a speech at Deutschlandfunk's conference "Formate des Politischen 2016": \url{https://vimeo.com/190347766}}. While he expressed his support for the decision to take this stance at the time, he argued that the German media is self-censoring and generally shows too little tolerance towards anti-immigration attitudes, thereby excluding those holding such views from political discourse.

This case presents a unique case to study the effects of media content on public opinion. As I will show below, the emphasis of crime strongly increased in *Bild*'s migration coverage following Reichelt's takeover. The change of editor itself was not discussed in the media, which means any opinion changes among readers can only be the result of the changing content [see @Spirig2020]. Additionally, the timing of the editorial change is potentially exogenous, given the pressure from allegations against Diekmann. In summary, *this case is as close as possible to a field experiment in which a specific consideration of migration coverage is amplified in the coverage of one newspaper, but not others*.

In the following sections, I will outline a framework for the measurement of emphasis frames in communication and showcase the increasing emphasis of crime in *Bild*'s migration coverage (the treatment). Subsequent sections will describe the panel data and variables to measure frames in thought (namely migration attitudes and their correlation with attitudes about crime) and define the difference-in-difference estimator. I will close with a discussion of the expected experimental effect size and minimal effects of interest to provide baselines for comparison of the estimator.

@@

To measure migration framing in the news, I collected 2.5 million newspaper articles from the websites of the most important German broadsheets **Frankfurter Allgemeine Zeitung (FAZ), Spiegel Online (SPON), Süddeutsche Zeitung (SZ), Die Tageszeitung (TAZ), Die Welt* and the major German tabloid *Bild* for the period 2013-2019 (for the market reach of each outlet, see table 1 in the appendix). Following the value-expectancy model, I identify crime frames in migration coverage in a two-step process: first, I use supervised machine learning to identify content about migration. Second, I identify content about crime *within* migration coverage by applying another supervised classifier. This approach closely resembles the value-expectancy model outlined in section \@ref(framing-effects), as I first identify coverage about the *issue*, before identifying the considerations emphasized in the discussion of this issue.

To identify whether an article is about migration, I employ supervised machine learning. As migration content is rare, and sufficient training data needs to be available to train functional classifiers, articles cannot be selected randomly for annotation\footnote{If migration content made up about 3% of all articles and the sample were selected randomly, one would find around 30 migration articles per 1000 articles, which provides rather limited information about the class of interest and results in worse recall of the classifier, as the best guess in highly imbalanced data is simply the most common outcome.}. Instead, I first construct a dictionary of terms related to migration. I use dictionary extension based on German GloVe word-embeddings\footnote{Downloadable from \url{https://deepset.ai/german-word-embeddings}.} to construct a comprehensive dictionary, and apply it to the articles. Based on the relative share of migration words in an article, I draw a stratified sample of 300 articles (100 from articles with no migration terms, 100 from the quarter with the highest share of migration terms and 100 from the remaining articles with relative frequencies of migration-related terms in between these two groups) for each newspaper for a total of 1,800 papers. Then, a coder hand-coded these articles, assessing whether their main topic is related to migration\footnote{A precise description of the classification process for migration and crime-related articles, including sampling dictionaries, instructions for coders, and preprocessing steps, can be found in appendix 1.2.}.

@@

TreatmentTrendPlot(dpa_corrected = T) + 
  TreatmentDiDPlot(dpa_corrected = T)

@@

The following section discusses the results. I will first present the above-discussed estimate of the effect on immigration attitudes and the association of immigration with crime attitudes. Then, I will assess the impact across time and on related variables to assess the robustness of the estimation. The last section will trace the key assumptions in the translation of experimental estimates discussed in section \@ref(framing-limits).
